---
title: "Grouped analysis across features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grouped analysis across features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(ptestR)
library(dplyr)
library(tidyr)
library(purrr)
```

## The real use case

In neuroimaging and multivariate biomedical research, it is common to run the
same model across many features simultaneously — brain states, EEG channels,
ROIs, metabolites. ptestR is designed for exactly this pattern.

This vignette demonstrates the `nest()` + `map()` workflow used in França et
al. (2024), where `grouped_perm_glmm()` was applied with 10 000 permutations
across dynamic functional connectivity features in 390 neonates from the
developing Human Connectome Project (dHCP).

---

## Simulated multi-feature dataset

We simulate a dataset resembling the dHCP structure: multiple subjects, each
with measurements across several brain states, and a set of clinical covariates.

```{r simulate}
set.seed(42)

n_subj   <- 40
n_states <- 6
features <- c("dwell_time", "transition_rate", "mean_amplitude")

covariates <- data.frame(
  subject        = seq_len(n_subj),
  PMA            = rnorm(n_subj, mean = 40, sd = 3),   # postmenstrual age
  PND            = rnorm(n_subj, mean = 10, sd = 5),   # postnatal days
  sex            = sample(c("M", "F"), n_subj, replace = TRUE),
  motion_outliers = runif(n_subj, 0, 0.15),
  group          = sample(c("term", "preterm"), n_subj, replace = TRUE)
)

# Expand to one row per subject × state × feature
df <- expand.grid(
  subject = seq_len(n_subj),
  state   = paste0("State_", seq_len(n_states)),
  feature = features,
  stringsAsFactors = FALSE
) |>
  left_join(covariates, by = "subject") |>
  mutate(
    value = rnorm(n()) +
            0.4 * (PMA - 40) * (feature == "dwell_time") +
            0.3 * (group == "preterm") * (feature == "mean_amplitude") +
            rnorm(n_subj, sd = 0.8)[subject]
  )

glimpse(df)
```

---

## Running across states and features

The pattern is: `group_by()` the grouping variables → `nest()` the data →
`map()` the permutation function → `unnest()` the results.

```{r grouped-glmm}
results <- df |>
  group_by(state, feature) |>
  nest() |>
  mutate(
    perm = map(data, \(d) grouped_perm_glmm(
      tbl         = d,
      formla      = value ~ PMA + PND + sex + motion_outliers + (1 | subject),
      var_to_perm = "value",
      permNum     = 499,   # use 9999 for publication
      seed        = 42
    ))
  ) |>
  unnest(perm) |>
  select(-data)

results
```

---

## Filtering to the terms of interest

The result contains all fixed-effect terms. In practice you will filter to the
predictors of interest — here, postmenstrual age:

```{r filter-pma}
pma_results <- results |>
  filter(term == "PMA") |>
  arrange(p.perm)

pma_results
```

---

## FDR correction

With `n_states × n_features` models, apply FDR correction across the full
result table for each term:

```{r fdr}
pma_results <- pma_results |>
  mutate(p.fdr = p.adjust(p.perm, method = "fdr"))

pma_results |>
  select(state, feature, estimate, statistic, p.perm, p.fdr) |>
  arrange(p.fdr)
```

This matches the analysis strategy in França et al. (2024): permutation
p-values computed per model, FDR correction at α = 5% across the full
feature set.

---

## Handling `p.perm = 0`

When no permuted statistic is as extreme as the observed one, `p.perm = 0`.
This is not a true zero — it is an upper bound. Report as `p < 1/permNum`:

```{r zero-perm}
pma_results |>
  mutate(
    p.perm.label = ifelse(
      p.perm == 0,
      paste0("p < ", round(1 / 499, 4)),
      paste0("p = ", round(p.perm, 4))
    )
  ) |>
  select(state, feature, term, p.perm.label) |>
  filter(p.perm == 0)
```

With `permNum = 9999` (recommended for publication), the bound becomes
`p < 0.0001`.

---

## Plotting the results

A simple heatmap of p-values across states and features:

```{r heatmap, fig.width = 6, fig.height = 3.5}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  pma_results |>
    mutate(
      sig       = p.fdr < 0.05,
      log10p    = -log10(pmax(p.perm, 1 / 499))
    ) |>
    ggplot(aes(x = state, y = feature, fill = log10p)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    geom_text(aes(label = ifelse(sig, "*", "")),
              size = 5, colour = "white", vjust = 0.8) +
    scale_fill_gradient(
      low  = "#F1E2D1",
      high = "#810B38",
      name = expression(-log[10](p[perm]))
    ) +
    labs(
      title = "Permutation p-values: PMA effect",
      x     = "Brain state",
      y     = "Feature",
      caption = "* FDR < 5%"
    ) +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

---

## Using `grouped_perm_glm()` instead

The same workflow applies for GLMs without random effects — just swap the
function and drop the random-effects term from the formula:

```{r grouped-glm}
glm_results <- df |>
  group_by(state, feature) |>
  nest() |>
  mutate(
    perm = map(data, \(d) grouped_perm_glm(
      tbl         = d,
      formla      = value ~ PMA + PND + sex + motion_outliers,
      var_to_perm = "value",
      permNum     = 299,
      seed        = 42
    ))
  ) |>
  unnest(perm) |>
  select(-data)

glm_results |>
  filter(term == "PMA") |>
  arrange(p.perm)
```

---

## Tips for large-scale analyses

**Increase `permNum` for publication.** 499 is used here for speed; use
9 999 or more for final results. At 10 000 permutations, `p.perm = 0` reports
as `p < 0.0001`.

**Parallelise across features, not within.** Each `grouped_perm_glmm()` call
is single-threaded. For large feature sets, use `furrr::future_map()` as a
drop-in replacement for `purrr::map()`:

```{r parallel, eval = FALSE}
library(furrr)
plan(multisession, workers = 4)

results <- df |>
  group_by(state, feature) |>
  nest() |>
  mutate(
    perm = future_map(data, \(d) grouped_perm_glmm(
      d,
      formla      = value ~ PMA + PND + sex + motion_outliers + (1 | subject),
      var_to_perm = "value",
      permNum     = 9999,
      seed        = 42
    ), .options = furrr_options(seed = TRUE))
  ) |>
  unnest(perm) |>
  select(-data)
```

**Save intermediate results.** For very large jobs, save after each feature
group and resume if interrupted.

---

## Reference

> França LGS, Ciarrusta J, Gale-Grant O, et al. (2024). Neonatal brain
> dynamic functional connectivity in term and preterm infants and its
> association with early childhood neurodevelopment. *Nature Communications*,
> 15, 16. <https://doi.org/10.1038/s41467-023-44050-z>

---

## Further reading

- `vignette("getting-started")` — core concepts and `grouped_perm_glm()`
- `vignette("mixed-effects-models")` — `grouped_perm_glmm()` in depth
- `?grouped_perm_glmm`, `?grouped_perm_glm`
